---
title: "GEOG 5023 Final Project"
author: "Isaiah Lyons-Galante & Caleb Schmitz"
date: "2023-04-22"
version: 2
output: html_document
---

## Setup Code
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load needed libraries
library(rdhs)
library(tidyverse)
library(haven)
library(sf)
library(gstat)
library(spdep)
```

## Manual Working Directory Setup
1. Put this file in a folder called "code". This is your working directory (wd).
2. Ensure you have the folder in the wd called "exports"
3. Ensure you have a folder in the wd called "data"
4. Ensure you have the folder "SLHR7ADT" inside of data, unzipped
5. Ensure you have the folder "SLGE7AFL" inside of data, unzipped
6. Ensure you have the folders with the regional and national boundaries as written below
7. Run this code!

## Import Data
```{r}
# manually read in the stata file
sl19 <- read_dta('./data/SLHR7ADT/SLHR7AFL.DTA')
var_labels <- get_variable_labels(sl19)

# read in shapefiles of the country, regions, and survey clusters
# https://gis.stackexchange.com/questions/19064/opening-shapefile-in-r

# rad in the shapefile of the country
country <- read_sf(dsn = "./data/sdr_national_data_2023-04-17/shps", layer = "sdr_national_data_dhs_2019")
plot(country)

# read in the shapefile of regions
regions <- read_sf(dsn = "./data/sdr_subnational_boundaries_2023-04-17/shps", layer = "sdr_subnational_boundaries2")
plot(regions)

# read in shapefile of survey cluster coordinates
clusters <- read_sf(dsn = "./data/SLGE7AFL", layer = "SLGE7AFL")
plot(clusters)

# "hv001" in sl19 dataframe tells us which cluster number each household belongs to
# "DHSCLUST" in gps dataframe contains the cluster number
# "LATNUM" and "LONGNUM" contain the coordinates of each cluster
# we'll need to join these two dfs to create maps of the variables we're interested in
# https://sparkbyexamples.com/r-programming/how-to-do-left-join-in-r/
sl19gps <- merge(
      x=sl19, 
      y=clusters, 
      by.x="hv001", 
      by.y="DHSCLUST",
      all.x=TRUE)

# export useful data to exports folder
saveRDS(country, "./exports/country.rds")
saveRDS(regions, "./exports/regions.rds")
saveRDS(clusters, "./exports/clusters.rds")
saveRDS(var_labels, "./exports/var_labels.rds")
saveRDS(sl19, "./exports/sl19.rds")
saveRDS(sl19gps, "./exports/sl19gps.rds")

```

## Clean and re-export
```{r}
sl19gps <- readRDS("./exports/sl19gps.rds")
df <- sl19gps

# columns of interest
interestingColumns <- c("hv000","hv001","hv006","hv007","hv010","hv011","hv012","hv013","hv014","hv024","hv025","hv040","hv045c","hv201","hv204","hv205","hv206","hv207","hv208","hv209","hv210","hv211","hv212","hv213","hv214","hv215","hv216","hv217","hv219","hv220","hv221","hv226","hv227","hv230a","hv237","hv241","hv243a","hv243b","hv243c","hv243d","hv243e","hv244","hv245","hv246","hv246a","hv246b","hv246c","hv246d","hv246e","hv246f","hv247","hv270","hv271","hv270a","hv271a","hml1")
gpsColumns <- c("LATNUM", "LONGNUM", "geometry", "ALT_DEM", "DATUM", "DHSREGNA", "ADM1NAME", "DHSID")
keepColumns <- c(interestingColumns, gpsColumns)

# filter down columns
df <- df[,keepColumns]

# filter out 0 coordinates
df <- filter(df, LATNUM != 0)

# re-export
sl19clean <- df
saveRDS(sl19clean, "./exports/sl19clean.rds")
```

## Create Maps of Variables
```{r}
## TODO: plot different variables from the survey using the gps coordinates of each cluster

# Dropping all data with coordinates (0,0) 
plottable_df <- filter(sl19gps, LATNUM > 0)

# This is turning hv206 from"haven-labelled' class to numeric for compatability with geom_point function.
plottable_df$hv206 <- as.factor(plottable_df$hv206)

ggplot(data = regions) +
    geom_sf() +
    geom_point(data=plottable_df, aes(x=LONGNUM, y=LATNUM, color=hv206)) +
    scale_color_manual(values = c("1" = "yellow", "0" = "black"))+
    labs(color="Access to Electricity")+ 
    ggtitle('Household Access to Electricity in Sierra Leone') + 
    theme_minimal()+
    ggspatial::annotation_scale()+
    ggspatial::annotation_north_arrow(location = "tr", which_north = "true") +
    labs(y= "", x = "") +
    theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank())
    theme(plot.title = element_text(size = 20))
    
    # We have a bit of a problem in that each cluster represents several families, so most likely we are just seeing one family per cluster which isn't necessarily representative. Perhaps a spatial interpolation method map would give a better overall impression of electricity availability.
    
    plottable_df$hv271 <- as.numeric(plottable_df$hv271)
    
    ggplot(data = regions) +
    geom_sf() +
    geom_point(data=plottable_df, aes(x=LONGNUM, y=LATNUM, color=(hv271))) +
    labs(color="Wealth Level")+ 
    ggtitle('Household Wealth Distribution in Sierra Leone') + 
    theme_minimal()+
    ggspatial::annotation_scale()+
    ggspatial::annotation_north_arrow(location = "tr", which_north = "true") +
    labs(y= "", x = "") +
    theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank())
    theme(plot.title = element_text(size = 20))
    
```


## Clean up data

### Chop down to 56 variables of interest
```{r}
# read in shortened RDS
sl19 <- readRDS("./exports/sl19.rds")
var_lables <- readRDS("./exports/var_labels.rds")

# columns of interest
keep2names <- c("hv000","hv001","hv006","hv007","hv010","hv011","hv012","hv013","hv014","hv024","hv025","hv040","hv045c","hv201","hv204","hv205","hv206","hv207","hv208","hv209","hv210","hv211","hv212","hv213","hv214","hv215","hv216","hv217","hv219","hv220","hv221","hv226","hv227","hv230a","hv237","hv241","hv243a","hv243b","hv243c","hv243d","hv243e","hv244","hv245","hv246","hv246a","hv246b","hv246c","hv246d","hv246e","hv246f","hv247","hv270","hv271","hv270a","hv271a","hml1")

# filter
sl19keep <- sl19[,keep2names]
vars2keep <- var_labels[keep2names,]

library(tidyverse)
library(haven)
# convert certain columns to factors
factorCols <- c("hv000","hv001","hv006","hv007","hv024","hv025","hv045c","hv201","hv205","hv206","hv207","hv208","hv209","hv210","hv211","hv212","hv213","hv214","hv215","hv217","hv219","hv221","hv226","hv227","hv230a","hv237","hv241","hv243a","hv243b","hv243c","hv243d","hv243e","hv244","hv246","hv246a","hv246b","hv246c","hv246d","hv246e","hv246f","hv247","hv270","hv270a")
sl19keep <- sl19keep %>% mutate_at(factorCols, as.factor)

# export to RDS
saveRDS(vars2keep, "./exports/vars2keep.rds")
saveRDS(sl19keep, "./exports/sl19keep.rds")
```

## Summarize data

```{r}
# import data
sl19keep <- readRDS("./exports/sl19keep.rds")
n <- nrow(sl19keep)
```

Histograms
```{r}
# histograms

hist(sl19keep$hv040, main="Altitude (m)") # altitude in meters
plot(sl19keep$hv024, main="Region") # region
plot(sl19keep$hv045c, main="Native Language")

ilg_hist <- function(x, bins, label, fill) {
  return(ggplot() + 
      geom_histogram(aes(x), color="black", fill=fill, bins=bins) +
      xlab(label) +
      ylab("Count") +
      ggtitle(label) +
      theme_bw()
  )
}


household <- ilg_hist(sl19keep$hv012, 33, "# Household Members", "purple")
household
women <- ilg_hist(sl19keep$hv010, 11, "# Women aged 15-49", "purple")
women
men <- ilg_hist(sl19keep$hv011, 11, "# Men aged 15-49", "purple")
men
children <- ilg_hist(sl19keep$hv014, 10, "# Children under 5 y.o.", "purple")
children

altitude <- ilg_hist(sl19keep$hv040, 30, "Altitude", "blue")
altitude

timetowater <- ilg_hist(sl19keep$hv204, 30, "Time to Water Source", "darkgreen")
timetowater # filter out high "unknowns"
```

Bar Charts
```{r}

ilg_bar <- function(x, label, fill) {
  return(ggplot() + 
      geom_bar(aes(x), color="black", fill=fill) +
      xlab(label) +
      ylab("Count") +
      ggtitle(label) +
      theme_bw()
  )
}

region <- ilg_bar(sl19keep$hv024, "Region", "blue")
region

language <- ilg_bar(sl19keep$hv045c, "Language", "blue")
language

watersource <- ilg_bar(sl19keep$hv201, "Source of Drinking Water", "darkgreen")
watersource


```

Pie Charts
```{r}
# pie charts of asset ownership
elec <- sum(sl19keep$hv206 == "1")
radio <- sum(sl19keep$hv207 == "1")
tele <- sum(sl19keep$hv208 == "1")
fridge <- sum(sl19keep$hv209 == "1")
bicycle <- sum(sl19keep$hv210 == "1")
motorcycle <- sum(sl19keep$hv211 == "1")
car <- sum(sl19keep$hv212 == "1")

pie(c(elec, n-elec), labels = c("Yes", "No"), main="Have Electricity", col=c("purple","white"))
pie(c(radio, n-radio), labels = c("Yes", "No"), main="Have Radio", col=c("purple","white"))
pie(c(tele, n-tele), labels = c("Yes", "No"), main="Have Television", col=c("purple","white"))
pie(c(fridge, n-fridge), labels = c("Yes", "No"), main="Have Fridge", col=c("purple","white"))
pie(c(bicycle, n-bicycle), labels = c("Yes", "No"), main="Have Bicycle", col=c("purple","white"))
pie(c(motorcycle, n-motorcycle), labels = c("Yes", "No"), main="Have Motorcycle", col=c("purple","white"))
pie(c(car, n-car), labels = c("Yes", "No"), main="Have Car", col=c("purple","white"))
```

Multivariable Bar Chart
```{r}
# bar chart of asset ownership
counts <- c(elec,radio,tele,fridge,bicycle,motorcycle,car)
labels <- c("Electricity", "Radio", "Television", "Fridge", "Bicycle", "Motorcyle", "Car")
assets <- data.frame(asset=labels, count=counts)
ggplot(assets) +
  geom_bar(aes(y=count, x=asset), fill="purple", stat='identity') +
  ggtitle('Asset Ownership')+
  xlab('Asset') +
  ylab('Count') +
  ylim(0,15000) + 
  theme_bw() +
  scale_color_brewer(palette="Dark2")+
  geom_hline(yintercept=n, style="dashed", color="gray")
```



## Spatial Regression

```{r}
# reread in dataframe
sl19clean <- readRDS("./exports/sl19clean.rds")
df <- sl19clean

# define columns to keep
assetColumns <- c("hv206","hv207","hv208","hv209","hv210","hv211","hv212","hv221","hv227","hv243a","hv243b","hv243c","hv243d","hv243e","hv246a","hv246b","hv246c","hv246d","hv246e","hv246f","hv247", "hv271")
gpsColumns <- c("LATNUM", "LONGNUM", "ALT_DEM")
srColumns <- c(assetColumns, gpsColumns)


# filter big data frame to just columns needed for this
df <- df[,srColumns]

# convert all columns to numeric
df <- as.data.frame(lapply(df, as.numeric)) 

# add geometry back in
df$geometry <- sl19clean$geometry

# filter out "unknown" or "missing" values from survey
df <- na.omit(df) # remove NA values
df <- df %>% filter(
  hv246a < 95 &
  hv246b < 95 &
  hv246c < 95 &
  hv246d < 95 &
  hv246e < 95 &
  hv246f < 95
)

# inspect df to ensure ranges make sense
summary(df)

# remove geometry again
geometry <- df$geometry
df = select(df, -geometry)

# basic linear model
basic_linear_model <- lm(hv271 ~ ., data=df)
summary(basic_linear_model)

# map residuals
df$geometry <- geometry
df$residuals <- residuals(basic_linear_model)

# to do DEBUG
# ggplot(df) + 
#   geom_sf(aes(fill=residuals, geometry=geometry), color="transparent") +
#   scale_fill_gradient2() +
#   ggtitle("Map of Model Residuals")

# define neighors
# queen <- poly2nb(df)
# ny_qw <- nb2listw(queen)
# 
# lm.morantest(ny.lm, ny_qw)
```
## Convert point data into area data
```{r}
library(h3jsr)

df <- sl19clean

# put lat long columns first
df <- df %>% select(LATNUM, LONGNUM, everything())
df <- h3jsr::point_to_cell(df, res = 8, simple = FALSE)

```