---
title: "CSCI5622 Module 1"
author: "IsaiahLG"
date: "1/31/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load needed libraries
library(rdhs)
library(tidyverse)
```
## To Do List
- Add ordering to certain factor columns
- Filter out unknown values such as 999
- Pull old data from Sierra Leone
- Pull data from other countries

## Import Data

### Using the package rdhs
```{r}
# set credentials for accessing DHS API
set_rdhs_config(email = "isaiah.lyons-galante@colorado.edu",
                project = "Machine Learning Class Project")

# figure out what DHS uses for country IDs
ids <- dhs_countries(returnFields=c("CountryName", "DHS_CountryCode"))
str(ids)

# find all surveys for Sierra Leone
surveys <- dhs_surveys(countryIds = c("SL"),
                       surveyType = "DHS",
                       surveyYearStart = 2018)
head(surveys)
# desired survey ID = "SL2019DHS"

# find the datasets from this survey
datasets <- dhs_datasets(surveyIds = surveys$SurveyId,
                         fileFormat = "flat")
head(datasets, n=20)
# desired filename = SLHR7AFL.ZIP
```

### Download the data from DHS
```{r}
# download actual datasets
downloads <- get_datasets("SLHR7AFL.ZIP")
```

### Read in downloaded dataset, write to CSV
```{r}
# read in the dataset
sl19 <- readRDS(downloads$SLHR7AFL)
# head(sl19) # 3455 columns! Yikes. 

# get the variable labels for all column headers
var_labels <- get_variable_labels(sl19)
# head(var_labels) # 3455 rows now, just two columns, variable and description

# export data and labels to CSV for visual inspection and usability in other programs
saveRDS(var_labels, "./exports/var_labels.rds")
saveRDS(sl19, "./exports/sl19.rds")

# write to csv as well
# write.csv(var_labels, "./exports/var_labels.csv", row.names = FALSE)
# write.csv(sl19, "./exports/sl19.csv", row.names = FALSE)
```

## Clean up data

### Chop down dataset to just section 1 and convert some columns to factors
```{r}
# reread in data from csv files
sl19 <- readRDS("./exports/sl19.rds")
var_labels <- readRDS("./exports/var_labels.rds")

# drop all sections but the first one
sl19sec1 <- sl19[,c(1:148)] 
varSec1 <- var_labels[c(1:148),]

# export
saveRDS(sl19sec1, "./exports/sl19sec1.rds")
saveRDS(varSec1, "./exports/varSec1.rds")

# write to csv as well
# write.csv(varSec1, "./exports/varSec1.csv", row.names = FALSE)
# write.csv(sl19sec1, "./exports/sl19sec1.csv", row.names = FALSE)
```

### Chop down to 56 variables of interest
```{r}
# read in shortened RDS
sl19sec1 <- readRDS("./exports/sl19sec1.rds")
varSec1 <- readRDS("./exports/varSec1.rds")

# columns of interest
keep2names <- c("hv000","hv001","hv006","hv007","hv010","hv011","hv012","hv013","hv014","hv024","hv025","hv040","hv045c","hv201","hv204","hv205","hv206","hv207","hv208","hv209","hv210","hv211","hv212","hv213","hv214","hv215","hv216","hv217","hv219","hv220","hv221","hv226","hv227","hv230a","hv237","hv241","hv243a","hv243b","hv243c","hv243d","hv243e","hv244","hv245","hv246","hv246a","hv246b","hv246c","hv246d","hv246e","hv246f","hv247","hv270","hv271","hv270a","hv271a","hml1")

# filter
sl19keep <- sl19sec1[,keep2names]
vars2keep <- varSec1[keep2names,]

# convert certain columns to factors
factorCols <- c("hv000","hv001","hv006","hv007","hv024","hv025","hv045c","hv201","hv205","hv206","hv207","hv208","hv209","hv210","hv211","hv212","hv213","hv214","hv215","hv217","hv219","hv221","hv226","hv227","hv230a","hv237","hv241","hv243a","hv243b","hv243c","hv243d","hv243e","hv244","hv246","hv246a","hv246b","hv246c","hv246d","hv246e","hv246f","hv247","hv270","hv270a")
sl19keep <- sl19keep %>% mutate_at(factorCols, as.factor)

# export to RDS
saveRDS(vars2keep, "./exports/vars2keep.rds")
saveRDS(sl19keep, "./exports/sl19keep.rds")

# write to csv as well
# write.csv(varSec1, "./exports/varSec1.csv", row.names = FALSE)
# write.csv(sl19sec1, "./exports/sl19sec1.csv", row.names = FALSE)
```

## Summarize data

```{r}
# import data
sl19keep <- readRDS("./exports/sl19keep.rds")
n <- nrow(sl19keep)
```

Histograms
```{r}
# histograms

hist(sl19keep$hv040, main="Altitude (m)") # altitude in meters
plot(sl19keep$hv024, main="Region") # region
plot(sl19keep$hv045c, main="Native Language")

ilg_hist <- function(x, bins, label, fill) {
  return(ggplot() + 
      geom_histogram(aes(x), color="black", fill=fill, bins=bins) +
      xlab(label) +
      ylab("Count") +
      ggtitle(label) +
      theme_bw()
  )
}


household <- ilg_hist(sl19keep$hv012, 33, "# Household Members", "purple")
household
women <- ilg_hist(sl19keep$hv010, 11, "# Women aged 15-49", "purple")
women
men <- ilg_hist(sl19keep$hv011, 11, "# Men aged 15-49", "purple")
men
children <- ilg_hist(sl19keep$hv014, 10, "# Children under 5 y.o.", "purple")
children

altitude <- ilg_hist(sl19keep$hv040, 30, "Altitude", "blue")
altitude

timetowater <- ilg_hist(sl19keep$hv204, 30, "Time to Water Source", "darkgreen")
timetowater # filter out high "unknowns"
```

Bar Charts
```{r}

ilg_bar <- function(x, label, fill) {
  return(ggplot() + 
      geom_bar(aes(x), color="black", fill=fill) +
      xlab(label) +
      ylab("Count") +
      ggtitle(label) +
      theme_bw()
  )
}

region <- ilg_bar(sl19keep$hv024, "Region", "blue")
region

language <- ilg_bar(sl19keep$hv045c, "Language", "blue")
language

watersource <- ilg_bar(sl19keep$hv201, "Source of Drinking Water", "darkgreen")
watersource


```

Pie Charts
```{r}
# pie charts of asset ownership
elec <- sum(sl19keep$hv206 == "1")
radio <- sum(sl19keep$hv207 == "1")
tele <- sum(sl19keep$hv208 == "1")
fridge <- sum(sl19keep$hv209 == "1")
bicycle <- sum(sl19keep$hv210 == "1")
motorcycle <- sum(sl19keep$hv211 == "1")
car <- sum(sl19keep$hv212 == "1")

pie(c(elec, n-elec), labels = c("Yes", "No"), main="Have Electricity", col=c("purple","white"))
pie(c(radio, n-radio), labels = c("Yes", "No"), main="Have Radio", col=c("purple","white"))
pie(c(tele, n-tele), labels = c("Yes", "No"), main="Have Television", col=c("purple","white"))
pie(c(fridge, n-fridge), labels = c("Yes", "No"), main="Have Fridge", col=c("purple","white"))
pie(c(bicycle, n-bicycle), labels = c("Yes", "No"), main="Have Bicycle", col=c("purple","white"))
pie(c(motorcycle, n-motorcycle), labels = c("Yes", "No"), main="Have Motorcycle", col=c("purple","white"))
pie(c(car, n-car), labels = c("Yes", "No"), main="Have Car", col=c("purple","white"))
```

Multivariable Bar Chart
```{r}
# bar chart of asset ownership
counts <- c(elec,radio,tele,fridge,bicycle,motorcycle,car)
labels <- c("Electricity", "Radio", "Television", "Fridge", "Bicycle", "Motorcyle", "Car")
assets <- data.frame(asset=labels, count=counts)
ggplot(assets) +
  geom_bar(aes(y=count, x=asset), fill="purple", stat='identity') +
  ggtitle('Asset Ownership')+
  xlab('Asset') +
  ylab('Count') +
  ylim(0,15000) + 
  theme_bw() +
  scale_color_brewer(palette="Dark2")+
  geom_hline(yintercept=n, style="dashed", color="gray")
```
```{r}
library(haven)
df <- read_dta('./data/SLHR7ADT/SLHR7AFL.DTA')
```
